Methods for Analyzing Models of\\ Rod-Shaped Bacteria

### 2025-07-29 paper/main.tex
rename vertices from v to x

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{v}_i$.
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{v}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{v}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{v}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.8\textwidth]{overview.pdf}
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents (yellow, red) that interact with and depend on
        each other.
        Yellow blocks are provided by this work and red blocks require additional input such as
        modeling assumptions or data.
        Blue blocks indicate possible future extensions of this project.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices $\vec{v}_i$.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{v}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{v}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{v}_{i-1},\vec{v}_i$ can be used when modeling the
dynamics of these bacteria.

@@

% \begin{enumerate}
%     \item represent bacteria as polygon with vertices $\vec{v}_i$
%     \item Assumes that these vertices are not fundamental but rather a discretization of the actual
%         problem
% \end{enumerate}

@@

For visualization we use pyvista, which provides high-level wrappers around the popular VTK
library~\cite{vtkBook,Sullivan2019}.
The first stage of our visualization is to generate 3D meshes for each bacterium, which we consider
an intermediate representation.
We insert spheres of a given radius $r$ at each vertex $\vec{v}_i$ and connect adjacent vertices
$\vec{v}_i,\vec{v}_{i+1}$ with a cylinder, also of radius $r$ which is positioned such that its core
is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render of this result can be seen in figure~\ref{fig:progression-image-generation}.
Since we use a 3D rendering engine as an intermediate representation, the resulting image can have
overlaps of bacteria which are placed at higher or lower values along the z-direction.

@@

This section will explain how we can determine the collection of vertices $\vec{v}_i$ introduced in
subsection~\ref{subsection:mechanical-model-mechanics} from microscopic images
(Figure~\ref{fig:position-extraction-algorithm} (A)).
They will serve as initial, final and intermediate values with which we can compare our predictions.
The algorithms which we devised requires the input of cell-masks which assign a unique color for
every cell in the image.
Such masks can be generated by cell-segmentation tools or obtained by annottating microscopic images
by hand (Figure~\ref{fig:position-extraction-algorithm} (B)).
Afterwards, we proceed with the following 5 steps.
\begin{enumerate}
    \item Obtain submasks for each cell individually.
    \item Skeletonize submask with algorithm from Lee~\cite{Lee1994}
    \item Determine endpoints $\vec{q}_0,\vec{q}_1$.
        Assert that there are only 2 endpoints.
    \item Sort points such that they form a polygon.
    \item Interpolate vertices $\vec{v}_i$ from obtained polygon.
\end{enumerate}
In the first step, the image will be split into a collection of subimages that only contain
information for one single cell.
From there, we apply a skeletonization algorithm~\cite{Lee1994}
(Figure~\ref{fig:position-extraction-algorithm} (C)).
It has to be noted that there are multiple variants of skeletonization and in general, these classes
of algorithms can produce results which have intersection points and multiple endpoints.
This described behaviour is undesirable in our case.
We have found that for most practical examples, the algorithm devised by~\cite{Lee1994} has worked
well.\\
The skeletonization algorithm supplies us with a collection of unordered lattice points $\vec{p}_i$
(pixel coordinates) where the skeleton lives.
We identify the endpoints $\vec{q}_0,\vec{q}_1$ by counting the number of neighbors in the von
Neumann neighborhood for each lattice point.
Any point with exactly one neighboring point must be and endpoint.
If we encounter more or less than two points which fit this criterion, the routine can not commence
and is aborted.
Next, we start with one of the endpoints and determine the next neighbor which was not yet picked.
We list the points in this order, thus forming a sorted polygon of pixel coordinates.\\
In the last step, we take the sorted, extracted points $\vec{p}_i$ and interpolate $n$ values for
the desired number of vertices $\vec{v}_i$.

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/image001042.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.255\textwidth]
        {docs/source/_static/fitting-methods/algorithm/interpolate-positions.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}
    \caption{
        (A) Input image.
        (B) Result of the skeletonization algorithm for a single cell.
        (C) Vertices $\vec{v}_i$ which have been interpolated from the skeletonized cell.
        They represent the position of our cell-agents.
    }
    \label{fig:position-extraction-algorithm}
\end{figure}

@@

Note that the growth rate is proportional to the rod length $l$ which leads to exponential growth.
This process will increase the length of the cell indefenitely unless we impose a condition for the
division event.
We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods.
\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}
Afterwards, we start at one endpoint $\vec{v}_0$ and travel along the polygon defined by the
bacterial rod.
We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.

>>>>

%###################################################################################################
\section{Overview}
This work consists of many subcomponents which can be combined in order to achieve various tasks.
Figure~\ref{fig:flowchart-project-structure} shows a high-level overview of all components.
We assume that bacterial rods can be represented by a colleciton of vertices $\vec{x}_i$.
This assumption alone is enough to enable a cascade of other components.
Using the spatial representation, we provide visualization methods that create 3D models from the
given vertices which can then either be rendered or projected to a 2D plane.
The data extraction component determines the vertices $\vec{x}_i$ from given cell masks which
correspond to microscopic images.
This allows us in particular to turn any sequence of images (movie) into a sequence of bacterial
positions.
The dynamics of the vertices $\vec{x}_i$ are determined by a computational model which is specific
to the respective use-case and should be defined by the researcher.
It merely needs to utilize the aforementioned spatial representation of bacterial rods as a
collection of vertices $\vec{x}_i$.
This work will provide one example of how such a model can be defined.
Notably, the visualization and data extraction components do not require any knowledge about the
computational model, thus allowing us to reuse their functionality across multiple models.

@@

\begin{figure}[H]
    \centering
    % \includegraphics[width=0.8\textwidth]{overview.pdf}
    \includestandalone[width=0.8\textwidth]{figures/concept/overview}
    \caption{
        The project can be divided into subcomponents (yellow, red) that interact with and depend on
        each other.
        Yellow blocks are provided by this work and red blocks require additional input such as
        modeling assumptions or data.
        Blue blocks indicate possible future extensions of this project.
        Every component implicitly builds on the spatial representation of rod-shaped bacteria as
        vertices $\vec{x}_i$.
    }
    \label{fig:flowchart-project-structure}
\end{figure}

%###################################################################################################
\section{Methodology}
% --------------------------------------------------------------------------------------------------
\subsection{Spatial Representation of Rod-Shaped Bacteria}
This work makes use of a few but powerfull assumptions concerning the mathematical representation of
the cells in question.
In order to describe the spatial dynamics of individual agents, we need to define their position,
shape and how they interact with each other.
In order to provide as much flexibility as possible when constructing new models, we restrict
ourselves to assumptions concerning the position such that researchers can freely experiment with
various shapes and interaction types.
Furthermore, it is very important to strike a good balance between the restrictiveness of our
assumptions and generality in order to provide meaningful results and still be able to apply our
approach to a broader class of models.
Taking into account the natural shape of rod-like bacteria, it is clear that they can not be
represented with a singular point $\vec{x}$ in space.
We could extend this positional definition with a vector $\vec{w}$, indicating the direction of
extension but this would restrict ourselves to fully straight rod-shaped bacteria.
However, multiple experiments have shown that MReB is responsible for the growth and maintenance of
the elongated shape of bacteria~\cite{Erickson2001}.
By means of this protein, rod-shaped bacteria can bend elastically on short time-scales and
deform plastically over longer periods of time.
Our goal is to provide a model which can encapsulate these effects.
Thus, we extend our model from a single point to multiple vertices which are part of a 1-dimensional
polygon with two endpoints $\vec{x}_i$.
In essence, we discretize the (possibly non-straight) rod of the agent.
Figure~\ref{fig:mechanics-bacterium} shows a visualization of this approach.
The angles $\alpha_i$ between vertices $\vec{x}_{i-1},\vec{x}_i$ can be used when modeling the
dynamics of these bacteria.

@@

% \begin{enumerate}
%     \item represent bacteria as polygon with vertices $\vec{x}_i$
%     \item Assumes that these vertices are not fundamental but rather a discretization of the actual
%         problem
% \end{enumerate}

@@

For visualization we use pyvista, which provides high-level wrappers around the popular VTK
library~\cite{vtkBook,Sullivan2019}.
The first stage of our visualization is to generate 3D meshes for each bacterium, which we consider
an intermediate representation.
We insert spheres of a given radius $r$ at each vertex $\vec{x}_i$ and connect adjacent vertices
$\vec{x}_i,\vec{x}_{i+1}$ with a cylinder, also of radius $r$ which is positioned such that its core
is identical to the connecting line segment between the two vertices.
Finally, we combine the meshes of all of these objects to obtain a single mesh of one bacterium.
A render of this result can be seen in figure~\ref{fig:progression-image-generation}.
Since we use a 3D rendering engine as an intermediate representation, the resulting image can have
overlaps of bacteria which are placed at higher or lower values along the z-direction.

@@

This section will explain how we can determine the collection of vertices $\vec{x}_i$ introduced in
subsection~\ref{subsection:mechanical-model-mechanics} from microscopic images
(Figure~\ref{fig:position-extraction-algorithm} (A)).
They will serve as initial, final and intermediate values with which we can compare our predictions.
The algorithms which we devised requires the input of cell-masks which assign a unique color for
every cell in the image.
Such masks can be generated by cell-segmentation tools or obtained by annottating microscopic images
by hand (Figure~\ref{fig:position-extraction-algorithm} (B)).
Afterwards, we proceed with the following 5 steps.
\begin{enumerate}
    \item Obtain submasks for each cell individually.
    \item Skeletonize submask with algorithm from Lee~\cite{Lee1994}
    \item Determine endpoints $\vec{q}_0,\vec{q}_1$.
        Assert that there are only 2 endpoints.
    \item Sort points such that they form a polygon.
    \item Interpolate vertices $\vec{x}_i$ from obtained polygon.
\end{enumerate}
In the first step, the image will be split into a collection of subimages that only contain
information for one single cell.
From there, we apply a skeletonization algorithm~\cite{Lee1994}
(Figure~\ref{fig:position-extraction-algorithm} (C)).
It has to be noted that there are multiple variants of skeletonization and in general, these classes
of algorithms can produce results which have intersection points and multiple endpoints.
This described behaviour is undesirable in our case.
We have found that for most practical examples, the algorithm devised by~\cite{Lee1994} has worked
well.\\
The skeletonization algorithm supplies us with a collection of unordered lattice points $\vec{p}_i$
(pixel coordinates) where the skeleton lives.
We identify the endpoints $\vec{q}_0,\vec{q}_1$ by counting the number of neighbors in the von
Neumann neighborhood for each lattice point.
Any point with exactly one neighboring point must be and endpoint.
If we encounter more or less than two points which fit this criterion, the routine can not commence
and is aborted.
Next, we start with one of the endpoints and determine the next neighbor which was not yet picked.
We list the points in this order, thus forming a sorted polygon of pixel coordinates.\\
In the last step, we take the sorted, extracted points $\vec{p}_i$ and interpolate $n$ values for
the desired number of vertices $\vec{x}_i$.

\begin{figure}[H]
    \centering
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/image001042.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{A}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.36\textwidth]
        {docs/source/_static/fitting-methods/algorithm/mask-zoom.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{B}};
    \end{tikzonimage}
    \begin{tikzonimage}[width=0.255\textwidth]
        {docs/source/_static/fitting-methods/algorithm/interpolate-positions.png}
        \node at (0.025, 0.975)[anchor=north west, rectangle, draw, white, minimum width=15pt, minimum height=15pt]{\textbf{C}};
    \end{tikzonimage}
    \caption{
        (A) Input image.
        (B) Result of the skeletonization algorithm for a single cell.
        (C) Vertices $\vec{x}_i$ which have been interpolated from the skeletonized cell.
        They represent the position of our cell-agents.
    }
    \label{fig:position-extraction-algorithm}
\end{figure}

@@

Note that the growth rate is proportional to the rod length $l$ which leads to exponential growth.
This process will increase the length of the cell indefenitely unless we impose a condition for the
division event.
We define a fixed length threshold for the total length of the cell at which it divides.
To construct a new cell, we cannot simply copy the existing one twice, but we also need to adjust
internal parameters in the process.
The following actions need to be taken for the old and new agent.
We first calculate the new spring lengths of the two resulting rods.
\begin{equation}
    \tilde{l} = l_i\left(\frac{1}{2} - \frac{r}{\sum\limits_i l_i}\right)
    \label{eq:growth-rod-new-lengths}
\end{equation}
Afterwards, we start at one endpoint $\vec{x}_0$ and travel along the polygon defined by the
bacterial rod.
We stop when we have reached a length that matches the length of the new spring length $\tilde{l}$.
From here, we continue and iterate this procedure until we have found all new vertices $\vec{w}_i$.
Equation~\ref{eq:growth-rod-new-lengths} ensures that that the tips of the two new rods are touching
right after the division event.
